Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-10-29 paper/main.tex
reformulate interaction part

Calculating the force between cells requires performing a double integral over the length of the
cells.
To reduce the computational costs we use an approximative, simplified model.
Given a vertex $\vec{x}_i$ on one cell, we determine the closest point $\vec{p}$ on the polygonal
line given by the vertices $\{\vec{y}_j\}$ of the interacting cell.
Furthermore we determine the value $q\in[0,1]$ such that $\vec{p} = (1-q)\vec{y}_j + q\vec{y}_{j+1}$
for some specific $j$.
The force is then calculated between the points $\vec{x}_i$ and $\vec{p}$ and acts on
the vertex $\vec{y}_j,\vec{y}_{j+1}$ with relative strength $(1-q)$ and $q$.

@@

\begin{align}
    U(r) &= C\epsilon\left[ \left(\frac{\sigma}{r}\right)^n -
        \left(\frac{\sigma}{r}\right)^m\right]\\
    C &= \frac{n}{n-m}\left(\frac{n}{m}\right)^{\frac{n}{n-m}}
\end{align}

The strength of the potential is given by $\epsilon$ and the thickness of our rods
is encoded in $\sigma = (m/n)^{1/(n-m)}(R_1+R_2)$ where $R_1,R_2$ are the thicknesses (radii) of the
two interacting rod-like bacteria.
We utilize the shorthand notation $r=|\vec{x}_i - \vec{p}|$ for the distance between the two points
in question.
We numerically regularize the potential by imposing an upper bound $\beta$ on the norm of the
resulting force $\vec{F} = \vec{\nabla} V$.
Furthermore, we impose a cutoff parameter $\zeta$ which sets sets the force to zero when the
distance $r$ between the interacting points exceeds this threshold.
The resulting force is given by

\begin{equation}
    \vec{F}_{i,\text{external}} = \vec{F} \theta(r-\zeta) \min\left(1, \beta / |\vec{F}|\right)
\end{equation}

where $\theta$ is the heaviside step-function and the $\min$ term ensures the upper bound $\beta$.
A collection of configurations are displayed in figure~\ref{fig:model-bacterium} (B).

The physical interaction between the bacteria plays a crucial role for the overall dynamics of the
system.
By varying its parameters, the Mie potential is capable of describing a wide range of interaction
types from softer ones that permit some level of overlap to rigid ones with steep ascends for
distance valuels that correspond to overlaps.
In contrast, a popular alternative is the Morse potential~\cite{Morse1929}, which is less
customizable due to the lower amount of parameters and more restrictive mathematical formulation.
Therefore, the Mie potential is a good choice for a simple, interpretable interaction potential with
low amount of parameters that can be estimated by our data-driven approach.

@@

where $n_v$ is the number of vertices and $R$ the previously introduced thickness of the rod.
This relation ensures that that the tips of the two new rods are touching right after the division
event.
We subdivide the polygon given by the vertices of the mother cell in $n$ segments of length
$\tilde{l}$, one segment of length $2R$ and again $n$ segments of length $\tilde{l}$.
The first and latter groups determine the new vertices of the two daughter cells.
This approach assumes cells where the total length of the polygon is greater than its diameter
$n_vl>2R$.
We set the growth rates of the daughter cells explicitly during the division event.
This approach allows us to treat the growth rates of the daughter cells as full individual
parameters within our model.
In order to reuse this model, we also provide the option to assign growth rates by sampling from a
normal distribution with user-defined parameters.
Every other intracellular parameter is copied identically to the daughter cells.
Table~\ref{table:list-of-parameters} shows a list of all parameters used within our model.

\begin{table}[H]
    \centering
    \def\arraystretch{1.3}
    \begin{tabular}{l c c l c c}
        \textbf{Parameters} && \textbf{Units} & \textbf{Variables} && \textbf{Units}\\
        \toprule
        \multicolumn{4}{l}{\textbf{Mechanics}}\\
        \midrule
        Spring tension & $\gamma$ & $\unit{1\per\minute}$ & Position (vertices) & $\vec{x}_i$ & $\unit{\micro\metre}$\\
        Rigidity & $\eta$ & $\unit{\micro\metre\per\minute^2}$ & Velocity & $\partial_t\vec{x}_i$ & $\unit{\micro\metre\per\minute}$\\
        Damping & $\lambda$ & $\unit{\micro\metre}$ & Spring length & $l$ & $\unit{\micro\metre}$\\
        \multicolumn{4}{l}{\textbf{Interaction}}\\
        \midrule
        Thickness & $R$ & $\unit{\micro\metre}$\\
        Strength & $\epsilon$ & $\unit{\micro\metre^2\per\minute^2}$\\
        Exponent 1 & $n$ & $1$\\
        Exponent 2 & $m$ & $1$\\
        Bound & $\beta$ & $\unit{\micro\metre^2\per\minute^2}$\\
        Cutoff & $\zeta$ & $\unit{\micro\metre}$\\
        \multicolumn{4}{l}{\textbf{Growth \& Division}}\\
        \midrule
        Growth rate & $\mu$\\
        Spring length threshold & $l_\text{thresh}$\\
        \bottomrule
    \end{tabular}
    \caption{
        List of all parameters and variables.
    }
    \label{table:list-of-parameters}
\end{table}

>>>>

Calculating the force between cells requires performing a double integral over the two polygonal
paths of the cells.
To reduce the computational costs we use an approximative, simplified model.
Given a vertex $\vec{x}_i$ on one cell, we determine the closest point $\vec{p}$ on the polygonal
line given by the vertices $\{\vec{y}_j\}$ of the interacting cell.
Furthermore we determine the value $q\in[0,1]$ such that $\vec{p} = (1-q)\vec{y}_j + q\vec{y}_{j+1}$
for some specific $j$.
This approach is especially justified for short-ranged interactions.
The force is then calculated between the points $\vec{x}_i$ and $\vec{p}$ and acts on
the vertex $\vec{y}_j,\vec{y}_{j+1}$ with relative strength $(1-q)$ and $q$.

@@

\begin{align}
    V(r) &= V_0C\left[ \left(\frac{\sigma}{r}\right)^n -
        \left(\frac{\sigma}{r}\right)^m\right]\\
    C &= \frac{n}{n-m}\left(\frac{n}{m}\right)^{\frac{n}{n-m}}
\end{align}

The strength of the potential is given by $V_0$ and the thickness of our rods
is encoded in $\sigma = (m/n)^{1/(n-m)}(R_1+R_2)$ where $R_1,R_2$ are the thicknesses (radii) of the
two interacting rod-like bacteria.
We utilize the shorthand notation $r=|\vec{x}_i - \vec{p}|$ for the distance between the two points
in question.
We numerically regularize the potential by imposing an upper bound $\beta$ on the norm of the
resulting force $\vec{F} = -\vec{\nabla} V$.
Furthermore, we impose a cutoff parameter $\zeta$ which sets sets the force to zero when the
distance $r$ between the interacting points exceeds this threshold.
The resulting force is given by

\begin{equation}
    \vec{F}_{i,\text{external}} = \left\{
    \begin{array}{ll}
        \beta\vec{F}/|\vec{F}| & |\vec{F}| >= \beta\\
        \vec{0} & r > \zeta\\
        \vec{F} & \text{else}
    \end{array}\right.
    % \vec{F} \theta(r-\zeta) \min\left(1, \beta / |\vec{F}|\right)
\end{equation}

\textbf{@Christian Can you find a more elegant notation than this equation?}\\
where the first case ensures that the norm is bound and the second case ensures the cutoff.
A collection of configurations are displayed in figure~\ref{fig:model-bacterium} (B).

The physical interaction between the bacteria plays a crucial role for the overall dynamics of the
system.
By varying its parameters, the Mie potential is capable of describing a wide range of interaction
types from softer ones that permit some level of overlap to rigid ones with steep ascends for
distance valuels that correspond to overlaps.
In contrast, a popular alternative is the Morse potential~\cite{Morse1929}, which is less
customizable due to the lower amount of parameters and more restrictive mathematical formulation.
Therefore, the Mie potential is a good choice for a simple, interpretable interaction potential with
low amount of parameters that can be estimated.

@@

where $n_v$ is the number of vertices and $R$ the previously introduced thickness of the rod.
This relation ensures that that the tips of the two new rods are touching right after the division
event.
We subdivide the polygon given by the vertices of the mother cell in $n$ segments of length
$\tilde{l}$, one segment of length $2R$ and again $n$ segments of length $\tilde{l}$.
The first and latter groups determine the new vertices of the two daughter cells.
This approach assumes cells where the total length of the polygon is greater than its diameter
$n_vl>2R$.
We set the growth rates of the daughter cells explicitly during the division event.
This approach allows us to treat the growth rates of the daughter cells as full individual
parameters within our model.
In order to reuse this model, we also provide the option to assign growth rates by sampling from a
normal distribution with user-defined parameters.\textbf{@Christian Supplement?}
Every other intracellular parameter is copied identically to the daughter cells.
Table~\ref{table:list-of-parameters} shows a list of all parameters used within our model.

\begin{table}[H]
    \centering
    \def\arraystretch{1.3}
    \begin{tabular}{l c c l c c}
        \textbf{Parameters} && \textbf{Units} & \textbf{Variables} && \textbf{Units}\\
        \toprule
        \multicolumn{4}{l}{\textbf{Mechanics}}\\
        \midrule
        Spring tension & $\gamma$ & $\unit{1\per\minute}$ & Position (vertices) & $\vec{x}_i$ & $\unit{\micro\metre}$\\
        Rigidity & $\eta$ & $\unit{\micro\metre\per\minute^2}$ & Velocity & $\partial_t\vec{x}_i$ & $\unit{\micro\metre\per\minute}$\\
        Damping & $\lambda$ & $\unit{\micro\metre}$ & Spring length & $l$ & $\unit{\micro\metre}$\\
        \multicolumn{4}{l}{\textbf{Interaction}}\\
        \midrule
        Thickness & $R$ & $\unit{\micro\metre}$\\
        Strength & $V_0$ & $\unit{\micro\metre^2\per\minute^2}$\\
        Exponent 1 & $n$ & $1$\\
        Exponent 2 & $m$ & $1$\\
        Bound & $\beta$ & $\unit{\micro\metre^2\per\minute^2}$\\
        Cutoff & $\zeta$ & $\unit{\micro\metre}$\\
        \multicolumn{4}{l}{\textbf{Growth \& Division}}\\
        \midrule
        Growth rate & $\mu$\\
        Spring length threshold & $l_\text{thresh}$\\
        \bottomrule
    \end{tabular}
    \caption{
        List of all parameters and variables.
    }
    \label{table:list-of-parameters}
\end{table}
